---
title: "countries"
output: html_document
date: "2023-02-18"
---

# 1. Load and check data

``` {r, results="asis"}
library('tidyverse')
cntry <- read_delim("../Data/gapminder.csv")
```

2. (2pt) Load data. How many rows/columns do we have?

``` {r}
dim <- dim(cntry)
cat("there are", dim, "rows and columns, respectively")
knitr::kable(head(cntry))
```

# 2 Descriptive statistics

1. (3pt) How many countries are there in the dataset?

There are `r length(unique(cntry$name))` names of countries.
There are `r length(unique(cntry$iso2))` iso2 codes.
There are `r length(unique(cntry$iso3))` iso3 codes.

2. If you did this correctly, you saw that there are more names than iso-2 codes, and there are
even more iso3 -codes. What is going on? Can you find it out?

``` {r}
group_by(cntry, iso2) %>% 
  summarize(
    n = length(unique(name))
  ) %>% 
  filter(n > 1)

cat("There are two country names that do not have iso2 codes")

group_by(cntry, name) %>% 
  summarize(
    n = length(unique(iso3))
  ) %>% 
  filter(n > 1)

cntry[is.na(cntry$name),] %>% 
  group_by(iso3) %>% 
  summarize(
    n = n()
  )

cat("There are 4 unnamed countries with an iso3 code")
cat("These entities are CHANISL, GBM, KOS, and NLD_CURACAO")

```

3. (2pt) What is the minimum and maximum year in these data?

``` {r}
min(cntry[is.na(cntry$name),]$time)
max(cntry[is.na(cntry$name),]$time)
```

# 3 CO2 emissions 

1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r}
mssng <- group_by(cntry[is.na(cntry$co2) | is.na(cntry$co2_PC),], time) %>% 
  summarize(
    n = n()
  ) %>% 
  arrange(n)
head(mssng)

cat("2012, 2013, 2014 have the most missing data")

```

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. Add a few
more countries of your choice. Explain what do you see.
```{r}
plot <- filter(cntry, name == "United States of America" 
       | name == "India" 
       | name == "China"
       | name == "Korea, Republic of"
       | name == "Ukraine")

ggplot(data = plot) +
  geom_point(mapping = aes(x = time, y = co2)) +
  facet_wrap(~name)

cat("China has had the largest increase in co2 emissions. India, Korea, and the US also increased.")
cat("Ukraine is the only one of these countries that decreased in co2 emissions.")
```

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC). Make a similar plot of the
same countries. What does this figure suggest?
```{r}
ggplot(data = plot) +
  geom_point(mapping = aes(x = time, y = co2_PC)) +
  facet_wrap(~name)

cat("This shows that Korea, relative to how much people they have, actually increased their 
co2 emissions the most. Ukraine and the US decreased over time, while China and Indiaare steadily 
growing")
```

4. (6pt) Compute average CO2 emissions per capita across the continents (assume region is the
same as continent). Comment what do you see.
```{r}
co2plot <- cntry[!is.na(cntry$co2_PC) &
                 !is.na(cntry$region) &
                 (cntry$time == 1960 |
                 cntry$time == 2016),] %>% 
  group_by(region,time) %>% 
  summarize(
    avg = sum(co2_PC)/n()
  )

co2plot

## It seems as though countries with unnamed continents have the highest average co2 emmissions
## per capita. Aside from those countries, Asia, Europe and the Americas have very high co2 per
## capita, with Oceania being slightly less, and Africa by far having the least emissions. 
```

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita
across continents in 1960 and 2016.
```{r}
ggplot(co2plot, aes(fill=time, y=avg, x=region, group=avg)) +
  geom_bar(stat = "identity", position = "dodge")
```

6. Which countries are the three largest, and three smallest CO2 emitters (in terms of CO2 per
capita) in 2019 for each continent? (Assume region is continent).
  NOTE: I will find the 3 largest and smallest CO2 emitters in 2016 because no data was 
  collected
```{r}
## africa
afr <- cntry[!is.na(cntry$name) &
      !is.na(cntry$time) &
      !is.na(cntry$co2_PC) &
      cntry$region == "Africa" &
      cntry$time == "2016",] %>% 
  arrange(co2_PC) 
head(afr, 3)
tail(afr, 3)

cat("Africa's 3 smallest emitters and COD, SOM, and BDI, while the 3 largest were
SYC, LBY, and ZAF.")

## americas
amr <- cntry[!is.na(cntry$name) &
      !is.na(cntry$time) &
      !is.na(cntry$co2_PC) &
      cntry$region == "Americas" &
      cntry$time == "2016",] %>% 
  arrange(co2_PC) 
head(amr, 3)
tail(amr, 3)

cat("Americas' 3 smallest emitters and HTI, NIC, and HND, while the 3 largest were
CAN, USA, and TTO.")

## asia
asi <- cntry[!is.na(cntry$name) &
      !is.na(cntry$time) &
      !is.na(cntry$co2_PC) &
      cntry$region == "Asia" &
      cntry$time == "2016",] %>% 
  arrange(co2_PC) 
head(asi, 3)
tail(asi, 3)

cat("Asia's 3 smallest emitters and AFG, NPL, and YEM, while the 3 largest were
BHR, KWT, and QAT.")

## europe
eur <- cntry[!is.na(cntry$name) &
      !is.na(cntry$time) &
      !is.na(cntry$co2_PC) &
      cntry$region == "Europe" &
      cntry$time == "2016",] %>% 
  arrange(co2_PC) 
head(eur, 3)
tail(eur, 3)

cat("Europe's 3 smallest emitters and LIE, ALB, and MDA, while the 3 largest were
FRO, LUX, and GIB")

# oceania
oce <- cntry[!is.na(cntry$name) &
      !is.na(cntry$time) &
      !is.na(cntry$co2_PC) &
      cntry$region == "Oceania" &
      cntry$time == "2016",] %>% 
  arrange(co2_PC) 
head(oce, 3)
tail(oce, 3)

cat("Oceania's 3 smallest emitters and SLB, VUT, and KIR, while the 3 largest were
PLW, AUS, and NCL")
```

# 4 GDP per capita 

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
```{r}
ggplot(data = cntry[cntry$time == 1960,]) +
  geom_point(mapping = aes(x= GDP_PC, y=lifeExpectancy, size = totalPopulation, color=region))
```

2. (4pt) Make a similar plot, but this time use 2019 data only
```{r}
ggplot(data = cntry[cntry$time == 2019,]) +
  geom_point(mapping = aes(x= GDP_PC, y=lifeExpectancy, size = totalPopulation, color=region))
```

3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?

>Many of the larger countries (bigger dots) have moved up within the past few decades, meaning
that their life expectancy is now far higher. We also see that Asia, Europe, and the Americas 
all generally increased their life expectancy, while Africa and Oceania still remain low.

4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
```{r}
LE <- cntry[!is.na(cntry$region) & 
            !is.na(cntry$lifeExpectancy) & 
            (cntry$time == 1960 | cntry$time == 2019),] %>% 
  group_by(region, time) %>% 
  summarize(
    avg = sum(lifeExpectancy)/n()
  )
LE
cat("This table shows that all regions actually increased their life expectancy")
```

5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
```{r}
group_by(LE, region) %>% 
  summarize(
    change = avg[time == 2019] - avg[time == 1960]
  ) %>% 
  arrange(change)

cat("This table shows that Europe's life expectancy increased the least, while Asia's 
increased the most.")
```

6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r}
cntry[!is.na(cntry$name) &
      !is.na(cntry$GDP_PC) &
      (cntry$time == 1960 | cntry$time == 2019),] %>% 
  ggplot(aes(x = GDP_PC)) +
  geom_histogram(position = "dodge") +
  facet_wrap(~time)
```

7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
```{r}
## 1960
le60 <- cntry[!is.na(cntry$name) &
              !is.na(cntry$lifeExpectancy) &
              (cntry$time == 1960),]

usrank60 <- rank(-(le60$lifeExpectancy), na.last=TRUE, ties.method = "first")[
  which(le60$name == "United States of America")
]

cat("The rank of the US in 1960 was", usrank60)

## 2019
le19 <- cntry[!is.na(cntry$name) &
              !is.na(cntry$lifeExpectancy) &
              (cntry$time == 2019),]

usrank19 <- rank(-(le19$lifeExpectancy), na.last=TRUE, ties.method = "first")[
  which(le19$name == "United States of America")
]

cat("The rank of the US in 2019 was", usrank19)
```

8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?

relative ranks divided by corresponding number of countries:

1960: `r usrank60/length(unique(le60$name))`

2019: `r usrank19/length(unique(le19$name))`

# Finally tell us how many hours did you spend on this PS.

This PS took me about 6 hours.